---
title: "Entropy and KL divergence"
date: 2020-07-03
author: Nick Griffiths
categories: Statistics
---

```{r include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

## Shannon Entropy

Shannon entropy is a property of a probability distribution.

For a discrete distribution with only one event P(x) = 1, entropy is zero:

```{r}
library(entropy)

entropy(1)
```

Entropy is proportional to the number of binary variables (bits) we need to capture all (equally likely) discrete outcomes:

```{r}
entropy(c("1" = 1,"0" = 1)) # one binary variable

entropy(c("11" = 1,"00" = 1, "10" = 1, "01" = 1)) # two variables

entropy(c("111" = 1, "110" = 1, "101" = 1, "100" = 1,
          "011" = 1, "010" = 1, "001" = 1, "000" = 1)) # 3 variables
```

If we have two events and weight one of them more, entropy decreases. This reflects the fact that as we increase the weight, we approach the single event distribution with no entropy:

```{r}
entropy(c(1,1))

entropy(c(1,3))

entropy(c(1,10))

entropy(c(1,10^6))
```

Similarly, for continuous distributions, the entropy increases with flatter, wider, higher variance distributions:

```{r}
entropy(dnorm(-10:10))

entropy(dnorm(-10:10, sd = 100))
```

## Kullback–Leibler divergence

KL divergence describes differences in two distributions based on their entropies. The KL divergence of a distribution with itself is 0:

```{r}
KL.empirical(c(1,1), c(1,1))
```

It is the entropy carried by the *additional* binary variables (bits) needed by the higher-entropy distribution to fully represent the outcomes:

```{r}
dist1 <- c("11" = 1,"00" = 1, "10" = 1, "01" = 1) # 2 binary variables
dist2 <- c("111" = 1, "110" = 1, "101" = 1, "100" = 1,
          "011" = 1, "010" = 1, "001" = 1, "000" = 1) # 3 binary variables

KL.empirical(dist1, dist2)

entropy(c("1" = 1, "0" = 1)) # entropy of one binary variable
```

## Cross Entropy

Cross entropy is closely related to KL divergence and also contrasts two distributions. It is just the KL divergence plus the entropy of the lower-entropy distribution.

When the higher-entropy distribution (d2 below) has equally-weighted outcomes, cross-entropy is the same as its entropy regardless of the other distribution:

```{r}
d1 <- c(1,3,2,1,1)

d2 <- c(1,1,1,1,1)

entropy(d1)
entropy(d2)

entropy(d1) + KL.empirical(d1,d2)
```

Otherwise the cross-entropy could be either higher or lower. For example, if the lower-entropy distribution removes a higher-weighted value,  cross-entropy exceeds entropy:

```{r}
d1 <- c(1,0,2,1,1)
entropy(d1)

d2 <- c(1,2,2,1,1)

entropy(d2)
entropy(d1) + KL.empirical(d1,d2)
```

If it removes a lower-weighted value, cross-entropy is lower than entropy:

```{r}
d1 <- c(1,2,2,0,1)
entropy(d1)

d2 <- c(1,2,2,1,1)

entropy(d2)
entropy(d1) + KL.empirical(d1,d2)
```

Finally, cross-entropy has an important connection with the likelihood. If d1 is a model we've built, and d2 is the process we are modeling, then the cross-entropy is a scaled equivalent of the likelihood function. We can do maximum likelihood estimation by minimizing the cross-entropy.
