---
title: "Different kinds of odds"
author: Nick Griffiths
date: 2020-03-06
draft: true
categories: Statistics
---

A three-level outcome is modeled using a proportional odds model. The odds ratio (between the lowest level, and the first two levels) is fixed at 2. What are the potential risks?

```{r}
library(tidyverse)

o_to_p <- function (x) {
  x / (x + 1)
}

p_to_o <- function(x) {
  (x) / (1 - x)
}

three_level_props <- function (odds1, odds2) {
  p1 <- o_to_p(odds1)
  p12 <- o_to_p(odds2)
  p2 <- p12 - p1
  p3 <- 1 - p1 - p2

  c(p1, p2, p3)
}

make_row <- function(odds1, c) {
  tibble(
    odds1 = odds1,
    prob = three_level_props(odds1, c * odds1),
    grp = c("p1", "p2", "p3")
  )
}

make_data <- function(c) {
  map_df(seq(.1, 10, .1), ~ make_row(., c))
}

make_plot <- function(c) {
  ggplot(make_data(c), aes(x = odds1, y = prob, fill = grp)) +
    geom_col()
}
```

- How do the odds of each category (versus all others) and log odds of each category behave when probability changes?


How do the odds ratio of the second category, and odds ratio of third category, compared to the first, behave when probability changes?

```{r}
general_odds <- function (odds1, odds21) {
  p1 <- o_to_p(odds1)
  p2 <- o_to_p(odds21) - p1
  p3 <- 1 - p1 - p2

  or2 <- p2 / p1
  or3 <- p3 / p1

  c(log(or2), log(or3))
}

make_row_gen <- function(odds1, c) {
  tibble(
    odds1 = odds1,
    or = general_odds(odds1, c * odds1),
    grp = c("or2", "or3")
  )
}

make_data_gen <- function(c) {
  map_df(seq(1, 10, .1), ~ make_row_gen(., c))
}

make_plot_gen <- function(c) {
  ggplot(make_data_gen(c), aes(x = log(odds1), y = or, fill = grp)) +
    geom_col()
}

make_plot_gen(2)
```

- from epi: how do categorical predictor ORs impact makeup by category within one level of an outcome?
