---
title: "Recipe for R data pipelines"
author: Nick Griffiths
date: 2020-04-09
draft: true
categories: programming
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE)
```

There are a lot of different aspects to processing a raw dataset for analysis. It helps to tackle things in order and keep similar steps together.

## Make data accessible (ASAP)

### 1. Rename columns

Make all names short and informative. With `tidyverse`, the convention is to use underscore names (`col_one`).

### 2. Join datasets

Merged data is much easier to work with than separate datasets. Do this as early as possible.

In the tidy framework, a dataset is a table where named columns represent variables and rows that represent individual observations of those variables.

So, datasets are defined by what is considered an "observation." This could be an event, a person, a date, or any other unit. Any data expressed with the same unit of observation should be combined into a unified dataset.

Before joining, it's often necessary to pivot the data.

At this point, the data may be worth saving as an intermediate file.

## Prep for analysis

### 3. Recode

The point of recoding is to put data into a format that can be used in calculations or analysis.

Some guidelines:

- The best format for data depends on its purpose (filter, math, viz, modeling)
- Format the same information consistently
- Combine columns when possible (e.g. combine dates and times)
- Group-wise information should be recycled through the group

The information should not change, so the variables themselves can be overwritten.

### 4. Perform calculations

Use the following framework:

```{r results="hide"}
suppressMessages(library(dplyr))

mtcars %>%
  group_by(vs) %>%
  mutate(newhp = mean(hp[gear == 3]))
```

Here, `hp[gear == 3]` operates over observations of `hp` when `gear` is 3, and `mutate` calculates each mean within the groups of `vs`.

The simple combination of `group_by`, `mutate`, and `[` can handle a surprisingly wide variety of tasks.

Write some tests to ensure the calculation is correct.

### 5. Rearrange

If there are columns that define the observation unit, put those first.

When groups of columns are often used together, keep them in order so they can be selected with `select(first:last)`.

## R tools

Examples with the `nycflights13` data

```{r}
library(nycflights13)
```

### Testing

Many testing packages are available, including `tinytest`, `assertr`, and `testthat`.

```{r}
library(tinytest)
library(checkmate) # extends tinytest

mean_delays <- flights %>%
  mutate(mean_del = mean(dep_delay, na.rm = T))

expect_double(mean_delays$mean_del, any.missing = F)
```

### Viewing data

Summary stats:

```{r}
print(skimr::skim(flights[1:3])) # Better summary (print not needed)
```

Head of all variables:

```{r}
tibble::glimpse(flights) # Better head
```

### Getting help

Docs and examples:

```{r}
?sum

example(sum)
```
